Dynamic Stochastic Approximation Monte Carlo

### 2018-07-23 papers/histogram/sad.tex
add red text describing our new idea and why it is so

\subsubsection{The ``interesting'' energy range}
Since SAD explores all energy states it needs to have a reasonable idea
of what energy range corresponds to the temperature range of interest,
defined by $T_{\min}<T<\infty$.
The
simulation is responsible for determining and updating this energy
range.
Given the true entropy $S(E)$, we can define the interesting energy
range as
\begin{align}
  E(T_{\min}) <E< E(T=\infty)
\end{align}
where $E(T)$ is the energy that maximizes $S-E/T$.  In the course of the
simulation this precise energy may be challenging to evaluate accurately.
In order to ensure that we sample this entire energy range adequately,
we define two energy limits:  a high energy $E_H$ and a low
energy $E_L$ to define the range over which the energy histogram
is made flat. $E_H$ is chosen to be the greatest energy that
maximized the entropy at any time during the simulation.  Similarly,
$E_L$ is the lowest energy that ever maximized $S(E)-E/T_{\min}$ during
the simulation.  This creates a ``ratcheting'' effect, in which $E_H$
may only increase, while $E_L$ may only decrease over the course of the
simulation, which results in a conservative estimate of the range of
energies that need be sampled.

>>>>

{\color{red}
While for SAMC the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SAD}}$ is
thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
$t_0$ should have dimensions of entropy.
We begin with an estimate of the average value of the entropy (relative
to the lowest entropy at $T_{\min}$).  If we assume a quadratic
dependence on energy, this is given by
\begin{align}
\langle S\rangle \approx \frac13 \frac{E({T=\infty}) - E(T_{\min})}{T_{\min}}
\end{align}
We approximate this energy difference by the energy difference
$E_H -E_L$ where these energies are defined below in the following
section.  Early in the simulation, when the energy range is still
expanding, this average entropy is a good choice
for $t_0$.  However, as the simulation progresses, our entropy factor
needs to scale with the total number of energy states that have been
found $N_S$, since the product $N_S\langle S\rangle$ corresponds to the
total change of entropy required, given that each of the $N_S$ states
will has a separate weight $w(E)$.  Finally, at \emph{long} times, after
all the energies have been found and the weights are being refined
we wish for a lower $t_0$ in order to reduce the fluctuations in the
weights that result from large values for $\gamma_t$.  We enable this
by tracking the last time we encountered a new energy $t_L$, and
gradually transition to lower $t_0$ (but still asymptotically scaling
as $\gamma_t \propto 1/t$ to ensure eventual convergence).  The SAD
expression for $\gamma_t$ which incorporates these ideas is:
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     \frac{E_{H}-E_{L}}{3T_{\text{min}}t}
    \frac{
      E_S^2 + E_S t + \left(t/t_L-1\right)t
    }{
      E_S^2 + t + \left(t/t_L-1\right)t
    }
\end{align}
This factor asymptotically has the same $1/t$ behavior as the original
SAMC algorithm;
however now every time a new energy is discovered during the course of
the simulation, the update factor experiences a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly. Figure~\ref{fig:gamma-vs-t} compares $\gamma_t$ for SAD, WL,
and SAMC.  The update factor
for WL remains high for many iterations, and then decreases very rapidly,
while the SAMC $\gamma_t$ remains constant before dropping as $1/t$.
The update factor for SAD in contrast fluctuates
dynamically around a value less than 1 for early MC moves
(while finding new energy states), before dropping in a \davidsays{similar
to SAD, but described after we see the plot}.
}

\subsubsection{The ``interesting'' energy range}\label{sec:EHL}
Since SAD explores all energy states it needs to have a reasonable idea
of what energy range corresponds to the temperature range of interest,
defined by $T_{\min}<T<\infty$.
The
simulation is responsible for determining and updating this energy
range.
Given the true entropy $S(E)$, we can define the interesting energy
range as
%\begin{align}
  $E(T_{\min}) <E< E(T=\infty)$
%\end{align}
where $E(T)$ is the energy that maximizes $S-E/T$.  In the course of the
simulation this precise energy may be challenging to evaluate accurately.
In order to ensure that we sample this entire energy range adequately,
we define two energy limits:  a high energy $E_H$ and a low
energy $E_L$ to define the range over which the energy histogram
is made flat. $E_H$ is chosen to be the greatest energy that
maximized the entropy at any time during the simulation.  Similarly,
$E_L$ is the lowest energy that ever maximized $S(E)-E/T_{\min}$ during
the simulation.  This creates a ``ratcheting'' effect, in which $E_H$
may only increase, while $E_L$ may only decrease over the course of the
simulation, which results in a conservative estimate of the range of
energies that need be sampled.
